A routing protocol for an urban IoT based on a smart street lighting wireless sensor network

### 2015-03-19 2014_ELSEVIER_CEE_SI/paper_cee_special_issue.tex
Versão aceita condicionalmente

\title{\sout{A smart street lighting system based on \rev{an} IPv6-enabled wireless sensor network} \rev{A routing protocol for an urban IoT based on a smart street lighting wireless sensor network}}

@@

\begin{abstract}
%% Text of abstract
In this paper, we propose a \rev{routing protocol} for a smart street lighting system based on IPv6-enabled wireless sensor network. We discuss the main drawbacks of \rev{the} RPL \rev{(IPv6 Routing Protocol for
Low-Power and Lossy Networks)} \sout{routing protocol} for P2P \rev{(point-to-point)} communication in large-scale 6LoWPAN networks. Then, we address such drawbacks by proposing a new routing algorithm, named \textit{GeoRank}, which integrates RPL protocol with \sout{GOAFR} the position-based routing protocol \rev{GOAFR (Greedy Other Adaptive Face Routing)}. Simulations were performed on networks extracted from real street maps. The obtained results show that \textit{GeoRank} finds shorter routes than RPL in high link density conditions and than GOAFR in low link density conditions. Thus, \textit{GeoRank} shows to be adaptive to variable link densities found in large-scale networks. Further, \textit{GeoRank} avoids the use of \sout{bandwidth-consuming} \rev{DAO (Destination Advertisement Object)} control messages required in RPL, while being more scalable in terms of memory usage than storing-mode RPL.
\end{abstract}

@@

\rev{With  the increase in world population and the growing urbanization worldwide, cities will need to evolve towards an intelligent infrastructure in order to provide to its citizens a better living environment. This will be achieved through the use of information and communication technology (ICT), with the goal to provide public services that are more accessible and more efficient. This vision is embodied in the ``smart city'' concept \citep{Smart_City_2011, Smart_City_2013}}. 
%\refs{(A Smart City and the Applications, 2011, Global Perspective of Smart Cities: A Survey, 2013).}}

\rev{According to several researchers, to enable the smart city it will be required an urban ``Internet of Things (IoT)'' infrastructure, where sensors and actuators distributed across the city will be interconnected by a pervasive network \citep{Smart_City_IOT_2013,Smart_City_IOT_2014,Smart_City_IOT_2014b,Smart_City_IOT_2014c,Smart_City_IOT_2014d,Smart_City_IOT_2014e}. 
%\refs{(Developing an IoT Smart City Framework, Internet of Things for Smart Cities, An Information Framework for Creating a Smart City Through Internet of Things, Padova Smart City: an Urban Internet of Things Experimentation, A Generic IoT Architecture for Smart Cities, ClouT: Cloud of Things for Empowering the Citizen Clout in Smart Cities.)} 
This urban ``IoT'' will enable a myriad of services and applications for the smart city, such as structural health monitoring, waste management, air quality monitoring, noise monitoring, traffic congestion avoidance, city energy consumption tracking, smart parking, smart lighting, charging infrastructure for electrical vehicles, etc\citep{Smart_City_2011, Smart_City_2013}.} % \refs{(Global Perspective of Smart Cities: A Survey, 2013, Smart City and the Applications).}}

%creation of bike lane network to promote emission-free transport, city-wide wiring and sensorization for environmental data collection,

\rev{One of such applications is smart street lighting. The street lighting system (SLS) is one of the most important assets for a city, due to its role as a public service for the citizens and also due its pervasiveness. However, SLS is very costly to operate, with a share of about 40\% of the total amount of electricity spent in a city \citep{Smart_City_Smart_Light_2013}. %\refs{(Smart Lighting solutions for Smart Cities, 2013)}. 
Therefore, making it more energy efficient and less costly for maintaining is very important for the cities.}

\sout{The street lighting system of a city plays an important role on its infrastructure. It is responsible of guaranteeing the society security and living during nighttime, traffic safety, and city appearance. Most commonly used lamp types in street lighting are the high pressure discharge lamps, e.g.: mercury vapor lamps, high-pressure sodium (HPS) lamps, and metal halide (MH) lamps %\citep{HID_Lamps_2010a,HID_Lamps_2010b}. 
However, with the efficiency and color quality of light emitting diodes (LEDs) steadily improving, LED lamps have been recently introduced as an alternative to replace the conventional lighting systems %\citep{LED_jou_2010a,LED_jou_2010b}.
}

\rev{The use of LED lamps in street lighting started to be reported in the last few years, and it is being considered a very good solution to \sout{replace the HID-based lighting systems} improve the energy efficiency of the system and also to reduce the maintenance cost, because of their long lifetime, high luminous efficiency \sout{in mesopic visual conditions,} and high color rendering index (CRI) \citep{CRI_2008,CRI_2010}.} \sout{Also, LED lamps may also be dimmed to reduce its power consumption.
Therefore, the presence of LED lamps in the modern street lighting systems brings a need for better management, in order to decrease the maintenance cost and power consumption (e.g.: by lamp dimming).} \rev{Also, the addition of communication capability to the SLS decreases the energy consumption and the maintenance cost, because it allows for monitoring the status information of each lamp (state, current, voltage and power measurements, etc.) and to have better control of the system (e.g.: lamp dimming). \sout{If the system is connected to a network, the lamp power can be monitored and the luminary can be quickly replaced when a failure occurs, because a decrease of active power consumption is observed. Not only a preventive maintenance of the system can be easily programmed, but also software updates can be done (which allows future enhancements and addition of new functions).}  \sout{Then, a street lighting system with communication capability is justified by its advantages.} Then, a smart SLS, based on LED lamps and a communication network, is an interesting application in a smart city context, not only by its clearly advantages but also because it can be one of the foundations for a urban ``IoT'' due to its pervasiveness across the city \citep{Smart_City_Smart_Light_2013}. %\refs{(Smart Lighting solutions for Smart Cities)}. 
Therefore, it can also be an enabler for the aforementioned applications in the smart city.}

@@

\rev{The main drawbacks of the PLC technology are noisy medium, high signal attenuation, susceptibility to interference from nearby devices, time-varying network topology, high complexity and poor scalability, despite its existing infrastructure and extensive coverage. The most suitable area for PLC seems to be the medium voltage distribution network, usually combined with wireless technologies in the low voltage distribution network through a hybrid architecture \citep{PLC_drawbacks}. Similarly, the scalability and reliability of cellular networks are not questionable, especially under low load traffic. However, the literature has already demonstrated that the limitations of the cellular networks for energy-constrained and/or processing-constrained devices are the tight synchronization requirement and high signalling overhead \citep{celullar_iot}. Furthermore, the dependency on the cellular carriers are usually not desirable in the smart cities/internet of things context. On the other hand, wireless sensor networks scalability is highly dependent on the routing algorithm performance. Therefore, designing a routing algorithm with high scalability and low overhead is a challenging requirement. It is important to note that we do not completely discard the PLC and cellular technologies for smart cities, but just reinforced that such technologies may be implemented as hybrid solutions, in order to develop redundant or complementary systems.}

% Recently, IETF has been standardized 6LoWPAN protocol (RFC4944) as an adaptation layer to enable IEEE 802.15.4-based low power and lossy networks (LLNs)

Recently, the 6LoWPAN protocol has been standardized by the IETF \rev{(Internet Engineering Task Force)} as an adaptation layer to enable IEEE 802.15.4-based low power and lossy networks (LLNs), such as wireless sensor networks, to reuse \rev{the} existing IPv6 protocol for \rev{the} networking layer \citep{RFC4944}. \rev{Also, the IETF has standardized RPL as the routing protocol for IP-based LLNs \citep{RFC6550}}. \rev{According to the literature, the main reasons to use IPv6 protocol for the LLNs include the connectivity through the internet to enable the IoT, the 128-bit addressing space that enables uniquely addressable devices (``things''), the network layered architecture and the interoperability \citep{IPv6_LLN_2010, IPv6_LLN_2011, Smart_City_IOT_2014b}.}

\sout{The IP protocol is an open, mature and field proven technology, with free and well documented standards know as RFCs, and thus it is very reasonable to reuse it for wireless sensor networking instead of creating new proprietary protocols.} \sout{However, 6LoWPAN protocol does not define the routing protocol, which shall be defined apart. Thus, IETF has standardized RPL as the routing protocol for IP-based LLNs}.

However, \sout{very recent empirical studies showed that} RPL has some drawbacks, as higher amount of control messages than data messages, mainly for constructing routes from the sink to the nodes (downward routes in RPL vocabulary) \citep{Heurtefeux_RPL_2013,P2P_analysis_2010,P2P-RPL_2011}. \sout{By the other hand, in} In a street light system, nodes are static and a routing protocol could benefit of this condition to find routes with less control overhead than RPL. The routing algorithms that use position to find routes in a communication network are known as geographic routing algorithms and a large body of literature has been devoted to the research on those algorithms in last few decades \citep{Karp_GPSR_2000, Position_Based_Routing_Ad_Hoc_2001, On_Optimal_Geographic_Routing_2007, Denardin2011}.

Our main contribution in this paper is to propose a modification into RPL protocol, in order to better support P2P communication. The proposed technique intend to integrate a geographic routing algorithm in the RPL protocol to reduce the amount of control message required for P2P communication. We further propose \rev{that such a routing protocol can be used in} a smart street lighting system based on IPv6-enabled wireless sensor network \rev{for supporting diverse applications in an urban IoT} \sout{which makes use of the new routing algorithm for message exchanging}. This paper is organized as follows: Section \ref{secRequirements} addresses the requirements of a wireless sensor network to implement an smart street lighting network. Also, we argue about the benefits of creating a network with such coverage for other public services. Section \ref{sec6LoWPAN} presents an overview of the 6LoWPAN protocol, describing succinctly its operating modes, routing strategies and main drawbacks. In section \ref{secGeographicRouting} we address the geographic routing technique, highlighting its pros and cons. In section \ref{secGeoRank}, the \textit{GeoRank} algorithm is presented, depicting its working principle. We evaluate, by simulation, the performance of the proposed algorithm in Section \ref{secSimulation}. Finally, we conclude this paper in Section \ref{secConclusion}.

\section{Requirements and benefits of a smart street lighting system}
\label{secRequirements}
\rev{A smart street lighting system based on a network that covers entire cities is a widespread concept in the literature \citep{Our_IAS_2009, WSN_Street_Lighting_2010_1, Cellular_Street_Lighting_2010_1, WSN_Street_Lighting_2010_2,Pantoni_2011, Pantoni_2013, daFonseca_2015}. However, the requirements for such network are not always clearly defined. Despite that, through a simple analysis it is clear that most of the messages in this system will be towards the sink node, generating a converging traffic. As the majority of the WSN protocols are designed for such traffic (e.g., the RPL protocol), we are led to think that there are few possible contributions that researchers can make to improve the network data flow. Therefore, it is important to express a more complete set of requirements that such a system demands for a communication network, stating the main goals of a smart street lighting system. Based on previous works cited above, we can assert that such goals are usually monitoring tasks, such as measuring current, voltage, energy consumption and state of the lamp/ballast. All this kind of message are sporadic or may be distributed in time to not overload the network, which has limited bandwidth due to the IEEE 802.15.4 standard. Thus, it is not possible to support the need for high performance P2P communication only by the requirements of a smart street lighting system.}

\rev{The implementation of a network solely for the purpose of monitoring the street lighting system also appears outside the context of smart cities and smart grid. Fortunately, such network has a huge potential for other applications due to its coverage. For instance, it is possible to use this network in order to improve the public transport system, as well as avoid traffic jams through a better control of the traffic lights. The public transport system may benefit from such a network if the bus stops and train stations exchange information with the buses/trains, which makes possible to inform users of this system the position and/or possible arrival time of the desired bus/train line. Furthermore, if the traffic lights and others means of traffic monitoring can communicate, it is possible to better coordinate vehicular traffic flow, even enabling the automatic warning of congestion for vehicles on certain roads \citep{SmartCity_ITS_2013}.}

\rev{These applications present different traffic patterns than a street lighting system, especially for not having the need to aggregate data in a sink node. The communication is mostly divergent, relying on P2P or multicast messages. Further, it is completely dynamic and application dependent how often and how many packets will be exchanged in a given moment or even in a given location. Therefore, a routing protocol capable of handling such heterogeneous scenario should take the fastest possible decisions with the least number of control messages, but also always looking for the shortest path. All of these requirements and challenges make this subject attractive to several researchers.}

@@

6LoWPAN is a protocol definition, standardized in \rev{Request for Comments (RFCs)} 4944 and 6282 \citep{RFC4944,RFC6282}, that works as an adaptation layer between the IPv6 network layer and the IEEE 802.15.4 MAC (Media Access Control) layer, as depicted in figure \ref{stack_6lowpan}. It is required because IEEE 802.15.4 standard, which defines MAC and PHY layers for low power and low data rate wireless networks, is able to carry packets with maximum size of 127 bytes, while IPv6 packets requires a MTU (minimum transmission unit) of 1280 bytes. Thus, the overall goal of 6LoWPAN is to allow IPv6 packets to be carried on top of IEEE 802.15.4-based wireless sensor networks, by defining fragmentation and compression mechanisms of the IPv6 packets and thus enabling IPv6-based wireless sensor networks \citep{Mulligan_6LoWPAN_2007}.

In addition to fragmentation and header compression, 6LoWPAN also supports mesh networks, as most wireless sensor networks are expected to require multi-hop communication. Two possible configurations for mesh networking are defined by 6LoWPAN: \textit{mesh under} and \textit{route over}. In mesh under, the mesh networking must be provided by the link layer and all IPv6 hosts are only one IP hop away from the border router that is connected to an external IP network (this router is known as 6LoWPAN Border Router (6LBR)). \rev{On} \sout{By} the other hand, in route over, the routing is done at the IP layer and hosts may be multiple IP hops away from a 6LBR. Thus, every router in the networks act as an IP router. However, although 6LoWPAN supports mesh networking, the decision of how mesh routing is actually performed is out of the scope of the 6LoWPAN specification.

@@

Nodes running RPL exchange control messages to setup and maintain the DODAG, which in the 6LoWPAN implementation are ICMPv6 messages. The 1 byte code in the ICMPv6 header is used to differentiate between different sub-type of control messages \citep{RFC4443}. The sub-type used to build the DODAG is called DAG Information Object (DIO) \citep{RFC6550}. All the nodes in the network regularly broadcast a DIO to its one-hop neighbors, indicating their activity and rank. The instant at which a node issues a DIO is governed by a Trickle timer \citep{trickle}, which is reset only when an inconsistency arises, i. e., very few DIOs are exchanged when the topology is stable. RPL \rev{provides} optimized paths to one or more LBRs with very little information stored in the routing table. The information stored is limited to its one-hop neighbors and their parents in the DODAGs. A parent of a node within a DODAG is one of the immediate successors of the node on a path towards the DODAG root.

@@

The RPL protocol has several possible drawbacks, depending on the application. Generally, we can point out three usual problems. First, one can easily realize that the P2P routes provided by RPL can be much larger than the shortest available route, causing high traffic on the DODAGs routes (worsening nearby root nodes), as discussed in \citep{P2P_analysis_2010} and \citep{P2P-RPL_2011}. Second, the DAO message exchange may lead to excessive packet collisions, mainly in scenarios with lots of link losses. Thus, the generated downward routes are proactive and initiated by the destination. Therefore, in order \rev{for} \sout{to} an actuator \rev{to} make a request to a sensor, the sensor should know that an actuator needs to communicate by point-to-point mode. Third, proactive routing also impacts on the size of the routing table, i.e, the more down routes exists, the greater the memory required to store those paths.

@@

Due to these not negligible drawbacks, it is wise to develop some work around able to improve the RPL protocol deficiencies, mainly that related with P2P messages. \rev{To this end, a reactive distance vector routing protocol called LOADng (Lightweight Ad hoc On-Demand - Next Generation (LOADng)) has been proposed \citep{LOADng_2012}. However, in recently published papers (\citep{Tripathi2014, Vucinic_2013}, LOADng was compared to RPL  and, according to the authors, LOADng may even have worse performance than RPL according to several metrics, including average delay, control overhead and maximum memory requirements. Therefore, in this paper, we follow another possible approach for routing. Such approach} \sout{One of the possible approaches} is the use of geographic routing techniques, which are routing algorithms known by its low overhead and high scalability features \citep{Karp_GPSR_2000, Position_Based_Routing_Ad_Hoc_2001, On_Optimal_Geographic_Routing_2007}.

@@

Mainly due to its greedy step, GOAFR algorithm is able to perform quite well in networks with high link density, discovering often optimal or near-optimal paths, in contrast to RPL, which is constrained to route only through the DODAGs and thus, being unable to find the optimal paths. \rev{On} \sout{By} the other hand, in networks with low link density, mainly around the critical density \citep{GOAFR_2008}, GOAFR has a performance loss and may usually find longer paths than RPL. Therefore, a hybrid approach that leverages the best characteristics of RPL and GOAFR according to the network link density is proposed, resulting in better performance than a single instance of each one would present. The new routing protocol is named \textit{GeoRank} to represent the integration of the geometric based behavior of GOAFR with the rank based behavior of RPL. This approach is mainly suitable for large scale \sout{and heterogeneous} networks \rev{with non-uniform link density}, such as a wireless network for street lighting nodes, where the link density varies from low to high depending on the presence of obstructions and buildings in different places of the network.

@@

%Será que não é necessário algumas definições? (não sei, pq a maioria do que precisaria definir já é definido no protocolo RPL). Existe o conceito de reliable nodes? (acredito que não)

%O algoritmo do CDF completo foi colocado textualmente no artigo original. Será que cabe mesmo utilizar pseudo-code para descrevê-lo?

@@

\STATE \textbf{2. Greedy forwarding mode:} If there are one or more neighbors in $N(u)$ closer to the destination than the current node holding the packet $u$, forward the packet to the neighbor $v \in N(u)$ with the greater progress towards the destination. Repeat this approach while such condition exists. If there are no neighbor in such position, store the node $u$ position in the packet and switch to GeoRank mode.
%the info of the algorithm change and

@@

In order to evaluate the proposed technique, simulations were performed in scenarios based on real street maps exported from the OpenStreetMap (OSM) data set, which is an open collaborative mapping project \citep{OSM}. \rev{The simulator, written in Julia, is based on an idealistic MAC/PHY layer with disc model for radio propagation}. The scenarios have been chosen to cover a wide range of possible street layouts \rev{inside a square region}. They have been populated with nodes with an average distance of $40m$, in order to simulate a real street lighting system, with network sizes ranging from 100 up to 400 nodes. A sample of the chosen scenarios are shown in figure \ref{sim_scenarios}.

For every scenario, there were generated $1000$ random source-destination pairs, giving a coverage of up to $10$\% of the possible source-destination combinations. \rev{For the RPL, the DAG is constructed as the Breadth First Search (BFS) tree for the topology, thus the expected transmission count (ETX) equals one for the preferred parent}. Also, every simulation was performed for $10$ random positions of the DODAG root in every scenario, which gives a good coverage of the \rev{root placements in a square region} \sout{$5$ possible root placements (i.e. center and top/bottom-right/left corners)}, \rev{and accounts for the unpredictability in the root location due to several factors, such as the availability of infrastructure (e.g. internet connection),  future network scaling, etc}. Then, for the each source-destination pair, the route was determined by each algorithm under test (RPL, GOAFR, \textit{GeoRank} and Dijkstra' algorithm for shortest path) and length of each route was measured in number of hops. The average length of the routes was computed and the results are shown in figure \ref{sim_results} \rev{with 95\% confidence intervals}.

%For RPL rank computation, it was used the Objective Function 0 as defined in RFC6552.
\begin{figure}
\centering
\ifdefined\TWOCOL
\includegraphics[width=0.45\textwidth]{figures/sample_scenarios.eps}
\else
\includegraphics[width=1\textwidth]{figures/sample_scenarios.eps}
\fi
\caption{Sample of simulated scenarios with nodes populated.}
\label{sim_scenarios}
\end{figure}

@@

\rev{For comparison of the scalability of \textit{GeoRank} and RPL, the number of entries in the routing table was recorded for each node. Then, the maximum value of the size of the routing table was computed over all the nodes, because it gives the worst case memory requirement (excluding the DODAG root). The results are shown in figure \ref{sim_results2}.}

>>>>

\title{A routing protocol for an urban IoT based on a smart street lighting wireless sensor network}

@@

\begin{abstract}
%% Text of abstract
In this paper, we propose a routing protocol for a smart street lighting system based on IPv6-enabled wireless sensor network. We discuss the main drawbacks of the RPL (IPv6 Routing Protocol for
Low-Power and Lossy Networks) for P2P (point-to-point) communication in large-scale 6LoWPAN networks. Then, we address such drawbacks by proposing a new routing algorithm, named \textit{GeoRank}, which integrates RPL protocol with the position-based routing protocol GOAFR (Greedy Other Adaptive Face Routing). Simulations were performed on networks extracted from real street maps. The obtained results show that \textit{GeoRank} finds shorter routes than RPL in high link density conditions and than GOAFR in low link density conditions. Thus, \textit{GeoRank} shows to be adaptive to variable link densities found in large-scale networks. Further, \textit{GeoRank} avoids the use of DAO (Destination Advertisement Object) control messages required in RPL, while being more scalable in terms of memory usage than storing-mode RPL.
\end{abstract}

@@

With  the increase in world population and the growing urbanization worldwide, cities will need to evolve towards an intelligent infrastructure in order to provide to its citizens a better living environment. This will be achieved through the use of information and communication technology (ICT), with the goal to provide public services that are more accessible and more efficient. This vision is embodied in the ``smart city'' concept \citep{Smart_City_2011, Smart_City_2013}.

According to several researchers, to enable the smart city it will be required an urban ``Internet of Things (IoT)'' infrastructure, where sensors and actuators distributed across the city will be interconnected by a pervasive network \citep{Smart_City_IOT_2013,Smart_City_IOT_2014,Smart_City_IOT_2014b,Smart_City_IOT_2014c,Smart_City_IOT_2014d,Smart_City_IOT_2014e}. 
This urban ``IoT'' will enable a myriad of services and applications for the smart city, such as structural health monitoring, waste management, air quality monitoring, noise monitoring, traffic congestion avoidance, city energy consumption tracking, smart parking, smart lighting, charging infrastructure for electrical vehicles, etc \citep{Smart_City_2011, Smart_City_2013}.

One of such applications is smart street lighting. The street lighting system (SLS) is one of the most important assets for a city, due to its role as a public service for the citizens and also due its pervasiveness. However, SLS is very costly to operate, with a share of about 40\% of the total amount of electricity spent in a city \citep{Smart_City_Smart_Light_2013}. Therefore, making it more energy efficient and less costly for maintaining is very important for the cities.

The use of LED lamps in street lighting started to be reported in the last few years, and it is being considered a very good solution to improve the energy efficiency of the system and also to reduce the maintenance cost, because of their long lifetime, high luminous efficiency and high color rendering index (CRI) \citep{CRI_2008,CRI_2010}. Also, the addition of communication capability to the SLS decreases the energy consumption and the maintenance cost, because it allows for monitoring the status information of each lamp (state, current, voltage and power measurements, etc.) and to have better control of the system (e.g.: lamp dimming). Then, a smart SLS, based on LED lamps and a communication network, is an interesting application in a smart city context, not only by its clearly advantages but also because it can be one of the foundations for a urban ``IoT'' due to its pervasiveness across the city \citep{Smart_City_Smart_Light_2013}. Therefore, it can also be an enabler for the aforementioned applications in the smart city.

@@

The main drawbacks of the PLC technology are noisy medium, high signal attenuation, susceptibility to interference from nearby devices, time-varying network topology, high complexity and poor scalability, despite its existing infrastructure and extensive coverage. The most suitable area for PLC seems to be the medium voltage distribution network, usually combined with wireless technologies in the low voltage distribution network through a hybrid architecture \citep{PLC_drawbacks}. Similarly, the scalability and reliability of cellular networks are not questionable, especially under low load traffic. However, the literature has already demonstrated that the limitations of the cellular networks for energy-constrained and/or processing-constrained devices are the tight synchronization requirement and high signalling overhead \citep{celullar_iot}. Furthermore, the dependency on the cellular carriers are usually not desirable in the smart cities/internet of things context. On the other hand, wireless sensor networks scalability is highly dependent on the routing algorithm performance. Therefore, designing a routing algorithm with high scalability and low overhead is a challenging requirement. It is important to note that we do not completely discard the PLC and cellular technologies for smart cities, but just reinforced that such technologies may be implemented as hybrid solutions, in order to develop redundant or complementary systems.

Recently, the 6LoWPAN protocol has been standardized by the IETF (Internet Engineering Task Force) as an adaptation layer to enable IEEE 802.15.4-based low power and lossy networks (LLNs), such as wireless sensor networks, to reuse the existing IPv6 protocol for the networking layer \citep{RFC4944}. Also, the IETF has standardized RPL as the routing protocol for IP-based LLNs \citep{RFC6550}. According to the literature, the main reasons to use IPv6 protocol for the LLNs include the connectivity through the internet to enable the IoT, the 128-bit addressing space that enables uniquely addressable devices (``things''), the network layered architecture and the interoperability \citep{IPv6_LLN_2010, IPv6_LLN_2011, Smart_City_IOT_2014b}.

However, RPL has some drawbacks, as higher amount of control messages than data messages, mainly for constructing routes from the sink to the nodes (downward routes in RPL vocabulary) \citep{Heurtefeux_RPL_2013,P2P_analysis_2010,P2P-RPL_2011}. In a street light system, nodes are static and a routing protocol could benefit of this condition to find routes with less control overhead than RPL. The routing algorithms that use position to find routes in a communication network are known as geographic routing algorithms and a large body of literature has been devoted to the research on those algorithms in last few decades \citep{Karp_GPSR_2000, Position_Based_Routing_Ad_Hoc_2001, On_Optimal_Geographic_Routing_2007, Denardin2011}.

Our main contribution in this paper is to propose a modification into RPL protocol, in order to better support P2P communication. The proposed technique intend to integrate a geographic routing algorithm in the RPL protocol to reduce the amount of control message required for P2P communication. We further propose that such a routing protocol can be used in a smart street lighting system based on IPv6-enabled wireless sensor network for supporting diverse applications in an urban IoT. This paper is organized as follows: Section \ref{secRequirements} addresses the requirements of a wireless sensor network to implement an smart street lighting network. Also, we argue about the benefits of creating a network with such coverage for other public services. Section \ref{sec6LoWPAN} presents an overview of the 6LoWPAN protocol, describing succinctly its operating modes, routing strategies and main drawbacks. In section \ref{secGeographicRouting} we address the geographic routing technique, highlighting its pros and cons. In section \ref{secGeoRank}, the \textit{GeoRank} algorithm is presented, depicting its working principle. We evaluate, by simulation, the performance of the proposed algorithm in Section \ref{secSimulation}. Finally, we conclude this paper in Section \ref{secConclusion}.

\section{Requirements and benefits of a smart street lighting system}
\label{secRequirements}
A smart street lighting system based on a network that covers entire cities is a widespread concept in the literature \citep{Our_IAS_2009, WSN_Street_Lighting_2010_1, Cellular_Street_Lighting_2010_1, WSN_Street_Lighting_2010_2,Pantoni_2011, Pantoni_2013, daFonseca_2015}. However, the requirements for such network are not always clearly defined. Despite that, through a simple analysis it is clear that most of the messages in this system will be towards the sink node, generating a converging traffic. As the majority of the WSN protocols are designed for such traffic (e.g., the RPL protocol), we are led to think that there are few possible contributions that researchers can make to improve the network data flow. Therefore, it is important to express a more complete set of requirements that such a system demands for a communication network, stating the main goals of a smart street lighting system. Based on previous works cited above, we can assert that such goals are usually monitoring tasks, such as measuring current, voltage, energy consumption and state of the lamp/ballast. All this kind of message are sporadic or may be distributed in time to not overload the network, which has limited bandwidth due to the IEEE 802.15.4 standard. Thus, it is not possible to support the need for high performance P2P communication only by the requirements of a smart street lighting system.

The implementation of a network solely for the purpose of monitoring the street lighting system also appears outside the context of smart cities and smart grid. Fortunately, such network has a huge potential for other applications due to its coverage. For instance, it is possible to use this network in order to improve the public transport system, as well as avoid traffic jams through a better control of the traffic lights. The public transport system may benefit from such a network if the bus stops and train stations exchange information with the buses/trains, which makes possible to inform users of this system the position and/or possible arrival time of the desired bus/train line. Furthermore, if the traffic lights and others means of traffic monitoring can communicate, it is possible to better coordinate vehicular traffic flow, even enabling the automatic warning of congestion for vehicles on certain roads \citep{SmartCity_ITS_2013}.

These applications present different traffic patterns than a street lighting system, especially for not having the need to aggregate data in a sink node. The communication is mostly divergent, relying on P2P or multicast messages. Further, it is completely dynamic and application dependent how often and how many packets will be exchanged in a given moment or even in a given location. Therefore, a routing protocol capable of handling such heterogeneous scenario should take the fastest possible decisions with the least number of control messages, but also always looking for the shortest path. All of these requirements and challenges make this subject attractive to several researchers.

@@

6LoWPAN is a protocol definition, standardized in Request for Comments (RFCs) 4944 and 6282 \citep{RFC4944,RFC6282}, that works as an adaptation layer between the IPv6 network layer and the IEEE 802.15.4 MAC (Media Access Control) layer, as depicted in figure \ref{stack_6lowpan}. It is required because IEEE 802.15.4 standard, which defines MAC and PHY layers for low power and low data rate wireless networks, is able to carry packets with maximum size of 127 bytes, while IPv6 packets requires a MTU (minimum transmission unit) of 1280 bytes. Thus, the overall goal of 6LoWPAN is to allow IPv6 packets to be carried on top of IEEE 802.15.4-based wireless sensor networks, by defining fragmentation and compression mechanisms of the IPv6 packets and thus enabling IPv6-based wireless sensor networks \citep{Mulligan_6LoWPAN_2007}.

In addition to fragmentation and header compression, 6LoWPAN also supports mesh networks, as most wireless sensor networks are expected to require multi-hop communication. Two possible configurations for mesh networking are defined by 6LoWPAN: \textit{mesh under} and \textit{route over}. In mesh under, the mesh networking must be provided by the link layer and all IPv6 hosts are only one IP hop away from the border router that is connected to an external IP network (this router is known as 6LoWPAN Border Router (6LBR)). On the other hand, in route over, the routing is done at the IP layer and hosts may be multiple IP hops away from a 6LBR. Thus, every router in the networks act as an IP router. However, although 6LoWPAN supports mesh networking, the decision of how mesh routing is actually performed is out of the scope of the 6LoWPAN specification.

@@

Nodes running RPL exchange control messages to setup and maintain the DODAG, which in the 6LoWPAN implementation are ICMPv6 messages. The 1 byte code in the ICMPv6 header is used to differentiate between different sub-type of control messages \citep{RFC4443}. The sub-type used to build the DODAG is called DAG Information Object (DIO) \citep{RFC6550}. All the nodes in the network regularly broadcast a DIO to its one-hop neighbors, indicating their activity and rank. The instant at which a node issues a DIO is governed by a Trickle timer \citep{trickle}, which is reset only when an inconsistency arises, i. e., very few DIOs are exchanged when the topology is stable. RPL provides optimized paths to one or more LBRs with very little information stored in the routing table. The information stored is limited to its one-hop neighbors and their parents in the DODAGs. A parent of a node within a DODAG is one of the immediate successors of the node on a path towards the DODAG root.

@@

The RPL protocol has several possible drawbacks, depending on the application. Generally, we can point out three usual problems. First, one can easily realize that the P2P routes provided by RPL can be much larger than the shortest available route, causing high traffic on the DODAGs routes (worsening nearby root nodes), as discussed in \citep{P2P_analysis_2010} and \citep{P2P-RPL_2011}. Second, the DAO message exchange may lead to excessive packet collisions, mainly in scenarios with lots of link losses. Thus, the generated downward routes are proactive and initiated by the destination. Therefore, in order for an actuator to make a request to a sensor, the sensor should know that an actuator needs to communicate by point-to-point mode. Third, proactive routing also impacts on the size of the routing table, i.e, the more down routes exists, the greater the memory required to store those paths.

@@

Due to these not negligible drawbacks, it is wise to develop some work around able to improve the RPL protocol deficiencies, mainly that related with P2P messages. To this end, a reactive distance vector routing protocol called LOADng (Lightweight Ad hoc On-Demand - Next Generation (LOADng)) has been proposed \citep{LOADng_2012}. However, in recently published papers (\citep{Tripathi2014, Vucinic_2013}, LOADng was compared to RPL  and, according to the authors, LOADng may even have worse performance than RPL according to several metrics, including average delay, control overhead and maximum memory requirements. Therefore, in this paper, we follow another possible approach for routing. Such approach is the use of geographic routing techniques, which are routing algorithms known by its low overhead and high scalability features \citep{Karp_GPSR_2000, Position_Based_Routing_Ad_Hoc_2001, On_Optimal_Geographic_Routing_2007}.

@@

Mainly due to its greedy step, GOAFR algorithm is able to perform quite well in networks with high link density, discovering often optimal or near-optimal paths, in contrast to RPL, which is constrained to route only through the DODAGs and thus, being unable to find the optimal paths. On the other hand, in networks with low link density, mainly around the critical density \citep{GOAFR_2008}, GOAFR has a performance loss and may usually find longer paths than RPL. Therefore, a hybrid approach that leverages the best characteristics of RPL and GOAFR according to the network link density is proposed, resulting in better performance than a single instance of each one would present. The new routing protocol is named \textit{GeoRank} to represent the integration of the geometric based behavior of GOAFR with the rank based behavior of RPL. This approach is mainly suitable for large scale networks with non-uniform link density, such as a wireless network for street lighting nodes, where the link density varies from low to high depending on the presence of obstructions and buildings in different places of the network.

@@

@@

\STATE \textbf{2. Greedy forwarding mode:} If there are one or more neighbors in $N(u)$ closer to the destination than the current node holding the packet $u$, forward the packet to the neighbor $v \in N(u)$ with the greater progress towards the destination. Repeat this approach while such condition exists. If there are no neighbor in such position, store the node $u$ position in the packet and switch to GeoRank mode.

@@

In order to evaluate the proposed technique, simulations were performed in scenarios based on real street maps exported from the OpenStreetMap (OSM) data set, which is an open collaborative mapping project \citep{OSM}. The simulator, written in Julia, is based on an idealistic MAC/PHY layer with disc model for radio propagation. The scenarios have been chosen to cover a wide range of possible street layouts inside a square region. They have been populated with nodes with an average distance of $40m$, in order to simulate a real street lighting system, with network sizes ranging from 100 up to 400 nodes. A sample of the chosen scenarios are shown in figure \ref{sim_scenarios}.

For every scenario, there were generated $1000$ random source-destination pairs, giving a coverage of up to $10$\% of the possible source-destination combinations. For the RPL, the DAG is constructed as the Breadth First Search (BFS) tree for the topology, thus the expected transmission count (ETX) equals one for the preferred parent. Also, every simulation was performed for $10$ random positions of the DODAG root in every scenario, which gives a good coverage of the root placements in a square region, and accounts for the unpredictability in the root location due to several factors, such as the availability of infrastructure (e.g. internet connection),  future network scaling, etc. Then, for the each source-destination pair, the route was determined by each algorithm under test (RPL, GOAFR, \textit{GeoRank} and Dijkstra' algorithm for shortest path) and length of each route was measured in number of hops. The average length of the routes was computed and the results are shown in figure \ref{sim_results} with 95\% confidence intervals.

\begin{figure}
\centering
\ifdefined\TWOCOL
\includegraphics[width=0.45\textwidth]{figures/sample_scenarios.eps}
\else
\includegraphics[width=1\textwidth]{figures/sample_scenarios.eps}
\fi
\caption{Sample of simulated scenarios with nodes populated.}
\label{sim_scenarios}
\end{figure}

@@

For comparison of the scalability of \textit{GeoRank} and RPL, the number of entries in the routing table was recorded for each node. Then, the maximum value of the size of the routing table was computed over all the nodes, because it gives the worst case memory requirement (excluding the DODAG root). The results are shown in figure \ref{sim_results2}.
